---
title: "MEFISTO application to lonigtudinal microbiome data"
author: "Britta Velten"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
---

```{r setup, warning=FALSE, message=FALSE}
library(MOFA2)
library(magrittr)
library(tidyverse)
library(cowplot)

# set output director for  figures
today <- Sys.Date()
figdir = paste0("figures_microbiome-", today, "/ECAM/")
if(!dir.exists(figdir)) dir.create(figdir)
knitr::opts_chunk$set(dev = c("png", "pdf"),
                      fig.path = figdir)
```

# Loading data
## Load trained model
Here we load the trained model, which was trained in `run_ECAM.py`. 
We include interpolated data for missing time points.
```{r}
mm <- load_model("out/ECAM_microbiome-seed_2020.hdf5", load_interpol_Z = TRUE)
mm
```

## Load meta data
This file contains metadata on the samples, such as the delivery mode and diet of the babies.
```{r}
meta <- read.csv("data/processed_data/sample_metadata.csv")
```

Merge meta data with sample meta data of object
```{r}
# summarise on month level
meta <- meta %>%
  separate(X.SampleID, into = c("study_id", "group", "time_id"), sep = "\\.") %>%
  select(group, delivery, diet, sex, antiexposedall, month)
meta %<>% mutate(sample = paste0(group, "-", month, ".0"))
meta <- meta %>% group_by(sample, group, month) %>%
  summarise(diet = unique(diet),
            sex = unique(sex),
            delivery = unique(delivery),
            antiexposedall = ifelse(any(antiexposedall == "y"), "y", "n")) %>%  
  ungroup()

# add information to samples_metadata of object
samples_metadata(mm) <- left_join(samples_metadata(mm), meta,
                                  by= c("group", "sample", "month"))

# fill in group-level meta data for missing time points (same mode of delivery, diet and sex)
samples_metadata(mm) <- samples_metadata(mm) %>%
  group_by(group) %>%
  mutate(delivery = unique(delivery[!is.na(delivery)]),
         sex = unique(sex[!is.na(sex)]),
         diet = unique(diet[!is.na(diet)])) %>% ungroup() 
```

# Variance explained
Let's take a look in which groups each factor explains variation
```{r}
plot_variance_explained(mm, split_by = "view", x = "group") +
  theme(axis.text.x = element_text(angle = 90))
```

# Factor along time
## Colored by devlivery mode
```{r}
df <- plot_factors_vs_cov(mm, color_by = "delivery", return_data = TRUE)
cols <- RColorBrewer::brewer.pal(8, "Dark2")
cols4delivery <- c("#e6ab02", "#d95f02")
names(cols4delivery) <- c("Cesarean", "Vaginal")
gg_delivery <- ggplot(df, aes(x=value.covariate, y = value.factor, color = color_by)) +
  geom_point(size=0.3, alpha =0.3) + facet_wrap(~factor) +
  stat_summary(aes(group = color_by, col = color_by), geom = "line")+
  scale_color_manual(values = cols4delivery) +
  xlab("month of life") + guides(col = FALSE) +
  theme_bw() + ylab("Factor value")
gg_delivery
```

## Colored by diet
```{r}
cols4diet <- c("#1f78b4", "#b2df8a")
names(cols4diet) <- c("fd", "bd")
df <- plot_factors_vs_cov(mm, color_by = "diet", return_data = TRUE)
gg_diet <- ggplot(df, aes(x=value.covariate, y = value.factor, color = color_by)) +
  geom_point(size=0.3, alpha =0.3) + facet_wrap(~factor)+
  stat_summary(aes(group = color_by, col = color_by), geom = "line") +
  scale_color_manual(values = cols4diet) +
  xlab("month of life") + guides(col = FALSE) +
  theme_bw() + ylab("Factor value")
gg_diet
```

Combine the two plots.
```{r}
gg_fac <- cowplot::plot_grid(gg_delivery, gg_diet,
                             ncol = 1, axis = "lr", align = "vh", labels = LETTERS[1:2])
```

# Clustering of babies per factor
Next we can look into the clustering of babies on each factor.
```{r}
# get metadata of babies for annotation of groups in heatmap
df_anno <- samples_metadata(mm) %>% as.tibble() %>% 
  group_by(group) %>%
  summarise(diet = unique(diet),
            sex = unique(sex),
            delivery = unique(delivery),
            antiexposedall = ifelse(any(antiexposedall == "y", na.rm = TRUE), "y", "n")) %>%
  select(group, delivery, diet, sex, antiexposedall) %>% #.[!duplicated(.),] %>% 
  ungroup() %>%
  column_to_rownames("group") %>% rename(antiexposed = antiexposedall)

# set colors
anno_col <- list(diet = cols4diet,
                 delivery = cols4delivery,
                 sex = c(Female = "pink", Male = "lightblue"),
                 antiexposed = c(y = 'black', n= "white"))

wd <- 3.5
gg1 <- plot_group_kernel(mm, factors = 1, show_rownames = FALSE, show_colnames = FALSE,
                  annotation_row = df_anno,
                  annotation_names_row = FALSE,
                  annotation_legend= FALSE, annotation_colors=anno_col,
                  treeheight_row =5, treeheight_col =5,
                  main = "", legend = FALSE, fontsize =9,
                  cellheight = wd, cellwidth = wd)
gg2 <- plot_group_kernel(mm, factors = 2, show_rownames = FALSE, show_colnames = FALSE,
                  annotation_row = df_anno,
                  annotation_names_row = FALSE,
                  annotation_legend= TRUE, annotation_colors=anno_col,
                  treeheight_row =5, treeheight_col =5,
                  main = "", legend = FALSE, fontsize =9,
                  cellheight = wd, cellwidth = wd)

# combine plots
gghm <- cowplot::plot_grid(gg1, gg2, nrow = 1,
                          scale = 0.9, rel_widths = c(0.75, 1))
gghm
```

Factor 1 shows a distinct cluster of babies that were all born by vaginal delivery.

# Smoothness and sharedness of factors
## Smoothness bars
```{r, fig.height=1}
gg_bar <- plot_smoothness(mm)
```

## Sharedness bars
```{r, fig.height=1}
gg_shared <- plot_sharedness(mm)
```

# Interpolation
We can use the model for interpolation to missing months. We do this here for a subset of babies
```{r interpol, fig.width=6, fig.height=3}
df_observed <- plot_factors_vs_cov(mm, covariate = "month", return_data = TRUE)
df <- get_interpolated_factors(mm, as.data.frame = TRUE)
df %<>% mutate(sd = sqrt(variance), ymin = mean -1.96 * sd, ymax = mean + 1.96 * sd)
df %<>% left_join(samples_metadata(mm), by = c("group", "month"))
df %<>% left_join(df_observed, by = c("sample", "group", "month" = "value.covariate", "factor"))
gg_interpol <- df %>% filter(factor == "Factor1",
                             group %in% c("C001", "C002", "C004", "C005",
                                          "C006", "C008", "C009", "C010", "C024", "C042")) %>%
  ggplot(aes(x=month, y = mean, col = group)) +
  geom_point(aes(x= month, y = value.factor, col = group, shape = antiexposedall), size = 1) +
  geom_line(aes(y=mean,  col = group)) +
  geom_ribbon(aes(ymin=ymin, ymax = ymax, fill = group),
              alpha = .2, col = "gray", size = 0.1) +
  facet_wrap(~ delivery + factor, ncol = 2) + theme_classic() +
  guides(col=FALSE, fill = FALSE) + ylab('Factor value')
gg_interpol
```

# Scatterplot
We can also look at the factor values on the sample level.
Here each dot correspond to one time-point-baby combination.
```{r scatter, fig.width=9, fig.height=3.5}
gg_scatter <- plot_grid(
plot_factors(mm, color_by = "delivery") +
  theme(legend.position = "top") +
  scale_fill_manual(values = cols4delivery),
plot_factors(mm, color_by = "diet") +
  theme(legend.position = "top") +
  scale_fill_manual(values = cols4diet),
plot_factors(mm, color_by = "month") +
  theme(legend.position = "top"),
nrow = 1, align = "h", axis = "tb")

gg_scatter
```

# Weights
Next we have a look at the microbial species associated with the two factors.
For this we have a look at the weights of each factor.

First we load the meta information on the features.
```{r}
taxonomy <- read.csv("data/processed_data/taxonomy.csv" )
taxa <- taxonomy$Taxon
names(taxa) <- taxonomy$Feature.ID
features_names(mm) <- list(microbiome = taxa[features_names(mm)[[1]]])
features_metadata(mm) <- separate(features_metadata(mm), col = feature, 
                                  sep = ";",
                                  into = c("kingdom", "phylum", "class",
                                           "order", "family", "genus", "species"),
                                  remove = FALSE)
```

Let's first have a look at the top positive and top negative species on factor 1.
```{r}
# top negative ones (more dominant in vaginal)
get_weights(mm, factors = 1, as.data.frame = TRUE) %>%
  arrange(value) %>% head(20) %>%
  separate(col = feature, sep = ";",
           into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
           remove = FALSE) %>% select(class, order, family, genus, species) 

# top positive ones  (more dominant in C-section)
get_weights(mm, factors = 1, as.data.frame = TRUE) %>%
  arrange(-value) %>% head(20) %>%
  separate(col = feature, sep = ";",
           into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
           remove = FALSE) %>% select(class, order, family, genus, species) 
```

We now summarise this on the genus level.
```{r w1}
df_weights1 <- get_weights(mm, as.data.frame = T, factors = 1) %>%
  separate(col = feature, sep = ";",
           into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
           remove = FALSE)

# filter out genera with missing genera information and shorten name
df_weights1 %<>% filter(!is.na(genus), genus != "g__")
df_weights1 %<>% mutate(genus = gsub("g__","", genus))
df_weights1 %<>% mutate(genus = ifelse(genus == " Pseudoramibacter_Eubacterium",
                                       "Pseudoramibacter_Eub.", genus))

# summarize by mean weights across all species in the genus and 
# filter to top 10 positive and negative ones
df_top <- df_weights1 %>% group_by(genus) %>% 
  summarize(mean_weight = mean(value)) %>%
  ungroup() %>% 
  # cesarean is linked to higher latent values
  mutate(type = ifelse(mean_weight > 0, "Cesarean", "Vaginal")) %>%
  group_by(type) %>%
  slice_max(abs(mean_weight), n= 10) %>% ungroup() %>% arrange(mean_weight) %>%
  mutate(genus = factor(genus, levels = .$genus))

# plot
gg_w1 <-  df_top %>% 
  ggplot(aes(x= genus, y = mean_weight, fill = type)) + geom_bar(stat="identity") +
  coord_flip() + theme_bw()  + scale_fill_manual(values = cols4delivery) +
  theme(legend.position = "top") + xlab("") + guides(fill = guide_legend(title="")) +
  geom_point(data = filter(df_weights1, genus %in% df_top$genus),
             aes(x = genus, y = value), inherit.aes = FALSE, alpha = 0.3)  +
  ylab("Weight (Factor 1)")
gg_w1
```

Let's do the same for factor 2.
```{r w2}
df_weights2 <- get_weights(mm, as.data.frame = T, factors = 2) %>%
  separate(col = feature, sep = ";",
           into = c("kingdom", "phylum", "class", "order",
                    "family", "genus", "species"),
           remove = FALSE)

# filter out genera with missing genera information
df_weights2 %<>% filter(!is.na(genus), genus != "g__")
df_weights2 %<>% mutate(genus = gsub("g__","", genus))

# summarize by mean weights across all species in the genus and 
# filter to top 10 postive and negative ones
df_top <- df_weights2 %>% group_by(genus) %>% 
  summarize(mean_weight = mean(value)) %>%
  ungroup() %>% 
  # bd is linked to higher latent values
  mutate(type = ifelse(mean_weight > 0, "bd", "fd")) %>% 
  group_by(type) %>%
  slice_max(abs(mean_weight), n= 10) %>% ungroup() %>% arrange(mean_weight) %>%
  mutate(genus = factor(genus, levels = .$genus))

# plot
gg_w2 <-  df_top %>% 
  ggplot(aes(x= genus, y = mean_weight, fill = type)) + geom_bar(stat="identity") +
  coord_flip() + theme_bw()  + scale_fill_manual(values = cols4diet) +
  theme(legend.position = "top") + xlab("") + guides(fill = guide_legend(title="")) +
  geom_point(data = filter(df_weights2, genus %in% df_top$genus),
             aes(x = genus, y = value), inherit.aes = FALSE, alpha = 0.3) +
  ylab("Weight (Factor 2)")
```


# Figure
For completeness, we here assemble the whole figure on this application.
```{r Figure3, fig.height=7, fig.width=14}
cowplot::plot_grid(cowplot::plot_grid(gg_fac, gg_scatter,
                                      labels = c("", "C"), rel_widths = c(1,1.5),
                                      align = "h", axis = "tb"),
                   cowplot::plot_grid(gghm, gg_w1, gg_w2,
                                      rel_widths = c(2,1,1), nrow = 1,
                                      labels = c("D", "E"), scale = 0.99), 
                   ncol =1, labels = c("", "C", "D"), align = "v",
                   axis = "lr", rel_heights = c(1.1,1))
```

```{r}
sessionInfo()
```
