---
title: "Factor stability"
author: "Britta Velten"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(MOFA2)
library(reshape2)
library(magrittr)
```

Compare the stability of the factors to downsampling on the microbiome data (pre-processing following Martino et al, Nat Biotech 2021). As this results in missing values and negative values comparison is limited to CTF, MOFA2 and MEFISTO. (PCA, NMF and ZIFA require non-missing values and the latter two positive data and would require a different preprocessing, e.g. variance stabilization of counts with library size correction).

```{r prep}
cols4diet <- c("#1f78b4", "#b2df8a")
names(cols4diet) <- c("fd", "bd")

cols4delivery <- c("#e6ab02", "#d95f02")
names(cols4delivery) <- c("Cesarean", "Vaginal")

cols4methods <- c("steelblue3",
                  "gray", "darkgoldenrod3")
names(cols4methods) <- c("MEFISTO",
                         "MOFA2", "CTF") 

# helper function to load factors and weights of different models
source("utils_stability.R")

# out directory with trained models
outdir <- "out/"

# set output directory for  figures
today <- Sys.Date()
figdir <- paste0("figures_microbiome-", today, "/stability/")
if(!dir.exists(figdir)) {
  dir.create(figdir)
}

knitr::opts_chunk$set(dev = c("png", "pdf"), fig.path = figdir)
```

```{r}
samples_meta <- read.csv("data/processed_data/samples_metadata.csv")
features_meta <- read.csv("data/processed_data/features_metadata.csv")
features_meta <- rename(features_meta, feature_id = SampleID)
samples_meta <- rename(samples_meta, sample_id = SampleID)

samples_present <- samples_meta %>%
  mutate(sample_id = paste(child_id, month, sep = "_")) %>%
  .$sample_id %>% unique()
```


# Load models
```{r load_models, eval = !file.exists(file.path(outdir, "df_Z.csv"))}
# MEFISTO
MEFSITOres <- load_files(outdir = outdir,
                         pattern = "^MEFISTO")
MEFSITOres$method <- "MEFISTO"

# MOFA
MOFAres <- load_files(outdir = outdir,
                      pattern = "^MOFA")
MOFAres$method <- "MOFA2"

# CTF
CTFres <- load_files_CTF(outdir = outdir)
CTFres$method <- "CTF"

# gather results from all methods
df_Z <- bind_rows(MOFAres, MEFSITOres, CTFres)

# check
stopifnot(!any(duplicated(df_Z)))

# Add meta data
df_Z <- left_join(df_Z, samples_meta, by = "sample_id")

# Get samples which are present in the full (non-masked data)
df_Z %<>% filter(sample_id %in% samples_present)
df_Z %<>% mutate(Factor1 = ifelse(is.na(Factor1), 0, Factor1))
df_Z %<>% mutate(Factor2 = ifelse(is.na(Factor2), 0, Factor2))

write.csv(df_Z, file = file.path(outdir, "df_Z.csv"))
```

```{r}
df_Z <- read.csv(file = file.path(outdir, "df_Z.csv"))
```

# Factor stability
```{r corZ, fig.width=7, fig.height=4}
Zmat1 <- acast(df_Z, sample_id ~ method + seed + miss_times,
               value.var = "Factor1")
df_corZ <- abs(cor(Zmat1)) %>%
  melt(value.name = "correlation") %>%
  separate(Var1,into = c("method1", "seed1", "time_miss1"), sep = "_")%>%
  separate(Var2,into = c("method2", "seed2", "time_miss2"), sep = "_") %>% 
  filter(method1 ==  method2, time_miss1 == 0, time_miss2 >0) %>%
  select(method = method1, time_miss = time_miss2, seed = seed2, correlation)%>%
  mutate(time_miss = as.numeric(time_miss), factor = "Factor 1")

Zmat2 <- acast(df_Z, sample_id ~ method + seed + miss_times, value.var = "Factor2")
df_corZ2 <- abs(cor(Zmat2)) %>% melt(value.name = "correlation") %>%
  separate(Var1,into = c("method1", "seed1", "time_miss1"), sep = "_")%>%
  separate(Var2,into = c("method2", "seed2", "time_miss2"), sep = "_") %>% 
  filter(method1 ==  method2, time_miss1 == 0, time_miss2 >0) %>%
  select(method = method1, time_miss = time_miss2, seed = seed2, correlation)%>%
  mutate(time_miss = as.numeric(time_miss), factor = "Factor 2")

df_corZ <- rbind(df_corZ, df_corZ2)
df_corZ %>%
ggplot(aes(x=time_miss, y = correlation, col = method)) +
  stat_summary(fun.data = "mean_se") + theme_bw() +
  ylab("Correlation to factor values in full model") +
  stat_summary(geom = "line", fun.data = "mean_se") +
  facet_wrap(~factor) + scale_color_manual(values = cols4methods) +
  xlab(paste0("# missing samples (out of ",  length(samples_present),")"))
```


```{r}
sessionInfo()
```

