---
title: "Microbiome application of MEFISTO"
author: "Britta Velten"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(MOFA2)
library(reshape2)
library(cowplot)
source("utils.R")
source("utils_microbiome.R")
library(magrittr)

cols4diet <- c("#1f78b4", "#b2df8a")
names(cols4diet) <- c("fd", "bd")

cols4delivery <- c("#e6ab02", "#d95f02")
names(cols4delivery) <- c("Cesarean", "Vaginal")

# set output director for  figures
today <- Sys.Date()
figdir = paste0("figures_microbiome-", today, "/")
if(!dir.exists(figdir)) dir.create(figdir)
knitr::opts_chunk$set(dev = c("png", "pdf"),
                      fig.path = figdir)
```

```{r}
datadir <- "data/processed_data/"
```


# Load data
```{r}
samples_meta <- read.csv(file.path(datadir, "samples_metadata.csv"))
features_meta <- read.csv(file.path(datadir, "features_metadata.csv"))
data <- read.csv(file.path(datadir, "rclr_mat_filtered.csv"),row.names = 1)

colnames(data) <- gsub("X10249", "10249", colnames(data))
data <- as.matrix(data)
df_data <- melt(data, varnames = c("feature_id", "sample_id")) %>%
  left_join(select(samples_meta, sample_id = SampleID, child_id, month),
            by = "sample_id")

# average per month (done internally in CTF)
df_data <- df_data %>%
  group_by(month, child_id, feature_id) %>%
  summarise(value = mean(value)) %>% ungroup()

# make child-level df on meta data
meta_child <- select(samples_meta, child_id, delivery, diet, sex)
meta_child <- meta_child[!duplicated(meta_child),]

# expand
df_data <- complete(df_data,expand(df_data, child_id, month, feature_id))

# fill in child-level information
df_data$view <- "microbiome"
df_data <- mutate(df_data, sample_id = paste(child_id, month, sep ="_"))
df_data <- df_data %>% left_join(meta_child, by = "child_id")

# turn zeros to NAs (no longer meaningful, following Martino et al, Nat Biotech 2021)
df_data <- mutate(df_data, value = ifelse(value == 0, NA, value))

# center features
df_data <- df_data %>% group_by(feature_id) %>%
  summarise(mean_value = mean(value, na.rm = TRUE)) %>% ungroup() %>%
  left_join(df_data, ., by = "feature_id") %>%
  mutate(value = value - mean_value) %>% 
  select(-mean_value)

# rename as required by MOFA/MEFISTO
df_data <- rename(df_data, feature = feature_id,
                  sample = sample_id, group = child_id)
head(df_data)
```


# Create and train MOFA object
```{r}
# create the MOFA object and add times as covariate
MOFAobject_untrained <- create_mofa(data = df_data)
MOFAobject_untrained <- set_covariates(MOFAobject_untrained, covariates = "month")
MOFAobject_untrained
```

```{r}
traindata_dir <- "data/input_data_rclr/"
```

```{r, eval = !dir.exists(traindata_dir)}
# optional: export data for training from python
dir.create(traindata_dir)
write.csv(df_data, file = file.path(traindata_dir, "microbiome_data.csv"))
export_data(model = MOFAobject_untrained, dir = traindata_dir)
write.csv(groups_names(MOFAobject_untrained),
          file = file.path(traindata_dir, "groups.csv"))
write.csv(views_names(MOFAobject_untrained),
          file = file.path(traindata_dir, "views.csv"))
```

**Prepare the MOFA object**
```{r MOFA_options}
data_opts <- get_default_data_options(MOFAobject_untrained)
data_opts$center_groups <- FALSE

model_opts <- get_default_model_options(MOFAobject_untrained)
model_opts$num_factors <- 2

mefisto_opts <- get_default_mefisto_options(MOFAobject_untrained)
mefisto_opts$n_grid <- 10
mefisto_opts$start_opt <- 50
mefisto_opts$opt_freq <- 50

train_opts <- get_default_training_options(MOFAobject_untrained)
train_opts$seed <- 2020
```

```{r prepare_MOFA, message = FALSE}
MOFAobject_untrained <- prepare_mofa(
  object = MOFAobject_untrained,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts,
  mefisto_options = mefisto_opts
) 
```

**Train the MOFA model**
```{r}
outdir <- "out/"
filename <- "MEFISTO_0_0.hdf5"
outfile <- file.path(outdir, filename)
```

```{r train_MOFA, eval = FALSE}
# MOFAobject <- run_mofa(MOFAobject_untrained, outfile = outfile)
```

For all further analysis we will use a pre-trained model. If this is not yet available, we can run `run_ECAM.py` (or `submit_training.sh`) to train the model from python on the data exported above from the untrained MOFA object. Once the trained models are obtained, we can continue with the downstream analysis below.
```{r}
knitr::opts_chunk$set(eval = file.exists(outfile), echo = file.exists(outfile))
```

# Load trained model
```{r load_model, warning = FALSE}
MOFAobject <- load_model(outfile, load_interpol_Z = TRUE)
MOFAobject

# add columns with nice name to metadata 
df <- samples_metadata(MOFAobject)
samples_metadata(MOFAobject) <- left_join(df, samples_metadata(MOFAobject_untrained),
                                          by = c("sample", "group","month"))
```

# Downstream analyses

## Variance explained by factor
Let's have a look how much variance the two factors explain jointly in the data.
```{r}
summary(Reduce(c,calculate_variance_explained(MOFAobject)$r2_total))
```


## Interpolation
Using the learnt Gaussian Processes, we can extrapolate or interpolate the factors to missing time intervals.
```{r, fig.width=12, fig.height=12}
plot_interpolation_vs_covariate(MOFAobject, only_mean = F) + facet_wrap(~group)
plot_interpolation_vs_covariate(MOFAobject, only_mean = F, factor = 2) + facet_wrap(~group)
```

## Factors versus month of life
Here, we plot the inferred factors versus month of life coloured by diet and delivery mode of the children.
```{r, warning=F, message=F}
median_IQR <- function(x) {
  data.frame(y = median(x), # Median
             ymin = quantile(x)[2], # 1st quartile
             ymax = quantile(x)[4])  # 3rd quartile
}

df <- plot_factors_vs_cov(MOFAobject, color_by = "delivery", return_data = TRUE)
gg_delivery <- ggplot(df, aes(x=value.covariate, y = value.factor, color = color_by)) +
  geom_point(size=0.3, alpha =0.3) + facet_wrap(~factor) +
  stat_summary(aes(group = color_by, col = color_by), geom = "line", fun.y = "median")+
  stat_summary(aes(group = color_by, fill = color_by),
               fun.data = "median_IQR", geom = "ribbon", alpha = 0.2, colour = NA) +
  scale_color_manual(values = cols4delivery) +
  scale_fill_manual(values = cols4delivery) +
  xlab("month of life") + guides(fill =FALSE, col = guide_legend(title = "delivery")) +
  theme_bw() + ylab("Factor value")
gg_delivery

df <- plot_factors_vs_cov(MOFAobject, color_by = "diet", return_data = TRUE)
gg_diet <- ggplot(df, aes(x=value.covariate, y = value.factor, color = color_by)) +
  geom_point(size=0.3, alpha =0.3) + facet_wrap(~factor) +
  stat_summary(aes(group = color_by, col = color_by), geom = "line", fun.y = "median", se = TRUE)+
  stat_summary(aes(group = color_by, fill = color_by),
               fun.data = "median_IQR", geom = "ribbon", alpha = 0.2, colour = NA) +
  scale_color_manual(values = cols4diet) +
  scale_fill_manual(values = cols4diet) +
  xlab("month of life") + guides(fill =FALSE, col = guide_legend(title = "diet")) +
  theme_bw() + ylab("Factor value")
gg_diet
```

Combine the two plots.
```{r}
gg_fac <- cowplot::plot_grid(gg_delivery, gg_diet,
                             ncol = 1, axis = "lr", align = "vh", labels = LETTERS[1:2])
```

## Scatterplot
We can also look at the factor values on the sample level.
Here each dot correspond to one time-point-baby combination.
```{r scatter, fig.width=9, fig.height=3.5}
gg_scatter <- plot_grid(
ggExtra::ggMarginal(plot_factors(MOFAobject, color_by = "delivery") +
  theme(legend.position = "top") +
  scale_fill_manual(values = cols4delivery),
  type = "boxplot", groupColour = TRUE, fill = cols4delivery, size = 10),
ggExtra::ggMarginal(plot_factors(MOFAobject, color_by = "diet") +
  theme(legend.position = "top") +
  scale_fill_manual(values = cols4diet),
  type = "boxplot", groupColour = TRUE, fill = cols4diet, size = 10),
plot_factors(MOFAobject, color_by = "month") +
  theme(legend.position = "top"),
nrow = 1, axis = "tb")


gg_scatter
```

## Weights
Next we have a look at the microbial species associated with the two factors.
For this we have a look at the weights of each factor.

First we load the meta information on the features.
```{r}
taxonomy <- read.csv("data/processed_data/features_metadata.csv" )
features_metadata(MOFAobject) <- features_metadata(MOFAobject) %>%
  left_join(taxonomy, by = c("feature" = "SampleID")) %>%
  separate(col = Taxon, sep = ";",into = c("kingdom", "phylum", "class",
                                           "order", "family", "genus", "species"),
                                  remove = FALSE)

# make feature names based on taxonomy
MOFAobject_fcodes <- MOFAobject # keep old  in alternative object for later analyses
features_names(MOFAobject) <- list(microbiome = features_metadata(MOFAobject)$Taxon)
```

### Factor 1
Let's first have a look at the top positive and top negative species on factor 1.
```{r}
# top negative ones (more dominant in vaginal)
get_weights(MOFAobject, factors = 1, as.data.frame = TRUE) %>%
  arrange(value) %>% head(20) %>%
  separate(col = feature, sep = ";",
           into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
           remove = FALSE) %>% select(class, order, family, genus, species, value) 

# top positive ones  (more dominant in C-section)
get_weights(MOFAobject, factors = 1, as.data.frame = TRUE) %>%
  arrange(-value) %>% head(20) %>%
  separate(col = feature, sep = ";",
           into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
           remove = FALSE) %>% select(class, order, family, genus, species) 
```

We now summarise this on the genus level.
```{r w1}
df_weights1 <- get_weights(MOFAobject, as.data.frame = T, factors = 1) %>%
  separate(col = feature, sep = ";",
           into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
           remove = FALSE)

gg_w1 <- make_genera_plot(df_weights1, factor_name = "Factor1")
```

### Factor 2
Let's do the same for factor 2.
```{r}
get_weights(MOFAobject, as.data.frame = T, factors = 2) %>%
  arrange(-abs(value)) %>% select(value, feature) %>% head(10)
```

```{r w2}
df_weights2 <- get_weights(MOFAobject, as.data.frame = T, factors = 2) %>%
  separate(col = feature, sep = ";",
           into = c("kingdom", "phylum", "class", "order",
                    "family", "genus", "species"),
           remove = FALSE)

gg_w2 <- make_genera_plot(df_weights2, factor_name = "Factor2", type = "diet")
```

## Clustering of babies per factor
Next we can look into the clustering of babies on each factor.
```{r}
# get metadata of babies for annotation of groups in heatmap
df_anno <- samples_metadata(MOFAobject) %>% as.tibble() %>% 
  group_by(group) %>%
  summarise(diet = unique(diet),
            sex = unique(sex),
            delivery = unique(delivery)) %>%
  select(group, delivery, diet, sex) %>%
  ungroup() %>%
  column_to_rownames("group")

# set colors
anno_col <- list(diet = cols4diet,
                 delivery = cols4delivery,
                 sex = c(Female = "pink", Male = "lightblue"))

wd <- 3.5
gg1 <- plot_group_kernel(MOFAobject, factors = 1, show_rownames = FALSE, show_colnames = FALSE,
                  annotation_row = df_anno,
                  annotation_names_row = FALSE,
                  annotation_legend= FALSE, annotation_colors=anno_col,
                  treeheight_row =5, treeheight_col =5,
                  main = "", legend = FALSE, fontsize =9,
                  cellheight = wd, cellwidth = wd)
gg2 <- plot_group_kernel(MOFAobject, factors = 2, show_rownames = FALSE, show_colnames = FALSE,
                  annotation_row = df_anno,
                  annotation_names_row = FALSE,
                  annotation_legend= TRUE, annotation_colors=anno_col,
                  treeheight_row =5, treeheight_col =5,
                  main = "", legend = FALSE, fontsize =9,
                  cellheight = wd, cellwidth = wd)

# combine plots
gghm <- cowplot::plot_grid(gg1, gg2, nrow = 1,
                          scale = 0.9, rel_widths = c(0.75, 1))
```


# Main figure
For completeness, we here assemble the whole figure on this application.
```{r Figure3, fig.height=7, fig.width=14}
cowplot::plot_grid(cowplot::plot_grid(gg_fac, gg_scatter,
                                      labels = c("", "C"),
                                      rel_widths = c(1,1.5),
                                      align = "h", axis = "tb"),
                   cowplot::plot_grid(gghm, gg_w1, gg_w2,
                                      rel_widths = c(2,1.2,1.2), nrow = 1,
                                      labels = c("D", "E"), scale = 0.99), 
                   ncol =1, labels = c("", "C", "D"), align = "v",
                   axis = "lr", rel_heights = c(1.1,1))
```


# Comparsion to CTF

## Factor 1
### CTF factor
```{r}
CTF_file <- "CTF_0_0_CTF_straj.csv"
factors_CTF <- read.csv(file.path(outdir,CTF_file))
factors_CTF<- factors_CTF %>% mutate(sample = paste(subject_id, month, sep= "_")) %>%
  select(sample, Factor1 = PC1, Factor2= PC2) %>%
  left_join(samples_metadata(MOFAobject), by = "sample")

gg_CTF_Z1 <- ggplot(factors_CTF, aes(x=month, y = Factor1, col = delivery)) +
  geom_point(alpha =0.2, size =0.5) + stat_summary(geom = "line") +
  scale_color_manual(values = cols4delivery) + theme_bw() +
  ggtitle("CTF component 1")
```

### CTF weights
```{r}
# get weights inferred by CTF
CTF_file <- "CTF_0_0_CTF_subj_ord_features.csv"
weights_CTF <- read.csv(file.path(outdir,CTF_file))
weights_CTF %<>% select(feature = X, Factor1 = PC2, Factor2 = PC2)

#scale to -1 to 1
weights_CTF$Factor1 <- weights_CTF$Factor1/ max(abs(weights_CTF$Factor1))
weights_CTF$Factor2 <- weights_CTF$Factor2/ max(abs(weights_CTF$Factor2))
weights_CTF %<>% gather(key = "factor", value = "value", -feature) 
weights_CTF %<>% left_join(taxonomy, by = c("feature" = "SampleID"))
weights_CTF %<>% separate(col = Taxon, sep = ";",
                           into = c("kingdom", "phylum", "class",
                                    "order", "family", "genus", "species"))
```

Make genera based plot for CTF
```{r}
df_CTF <- weights_CTF %>% filter(factor == "Factor1")

gg_w1_CTF <- make_genera_plot(df_CTF, factor_name = "Factor1", ngenera = 5) +
  ggtitle("Weights for CTF component 1")
gg_w1_CTF
```

### MEFISTO correspondences
```{r}
# factor
df <- plot_factors_vs_cov(MOFAobject, color_by = "delivery", return_data = TRUE)
gg_Z1 <- df %>% filter(factor == "Factor1") %>%
  ggplot(aes(x=value.covariate, y = value.factor, color = color_by)) +
  geom_point(size=0.3, alpha =0.3) + 
  stat_summary(aes(group = color_by, col = color_by), geom = "line")+
  scale_color_manual(values = cols4delivery) +
  xlab("month of life") + guides(col = FALSE) +
  theme_bw() + ylab("Factor value") + ggtitle("MEFISTO - Factor 1")

# weights
gg_w1 <- gg_w1 + ggtitle("Weights for MEFISTO Factor 1")
```

### Figure
```{r}
cowplot::plot_grid(gg_CTF_Z1 + guides(col=F) + xlab("month of life"),
  gg_w1_CTF + theme(legend.position = "right"),
  gg_Z1,
  gg_w1+ theme(legend.position = "right"),
  labels = LETTERS[1:4], rel_widths = c(1,1.5))
```

## Factor 2
### CTF factor
```{r}
CTF_file <- "CTF_0_0_CTF_straj.csv"
factors_CTF <- read.csv(file.path(outdir,CTF_file))
factors_CTF<- factors_CTF %>% mutate(sample = paste(subject_id, month, sep= "_")) %>%
  select(sample, Factor1 = PC1, Factor2= PC2) %>%
  left_join(samples_metadata(MOFAobject), by = "sample")

gg_CTF_Z2 <- ggplot(factors_CTF, aes(x=month, y = Factor2, col = delivery)) +
  geom_point(alpha =0.2, size =0.5) + stat_summary(geom = "line") +
  scale_color_manual(values = cols4delivery) + theme_bw() +
  ggtitle("CTF component 2")
```

### CTF weights
```{r}
# get weights inferred by CTF
CTF_file <- "CTF_0_0_CTF_subj_ord_features.csv"
weights_CTF <- read.csv(file.path(outdir,CTF_file))
weights_CTF %<>% select(feature = X, Factor1 = PC2, Factor2 = PC2)

#scale to -1 to 1
weights_CTF$Factor1 <- weights_CTF$Factor1/ max(abs(weights_CTF$Factor1))
weights_CTF$Factor2 <- weights_CTF$Factor2/ max(abs(weights_CTF$Factor2))
weights_CTF %<>% gather(key = "factor", value = "value", -feature) 
weights_CTF %<>% left_join(taxonomy, by = c("feature" = "SampleID"))
weights_CTF %<>% separate(col = Taxon, sep = ";",
                           into = c("kingdom", "phylum", "class",
                                    "order", "family", "genus", "species"))
```

Make genera based plot for CTF
```{r}
df_CTF <- weights_CTF %>% filter(factor == "Factor2")

gg_w2_CTF <- make_genera_plot(df_CTF, factor_name = "Factor2", ngenera = 5) +
  ggtitle("Weights for CTF component 2")
```

### MEFISTO correspondences
```{r}
# factor
df <- plot_factors_vs_cov(MOFAobject, color_by = "diet", return_data = TRUE)
gg_Z2 <- df %>% filter(factor == "Factor2") %>%
  ggplot(aes(x=value.covariate, y = value.factor, color = color_by)) +
  geom_point(size=0.3, alpha =0.3) + 
  stat_summary(aes(group = color_by, col = color_by), geom = "line")+
  scale_color_manual(values = cols4diet) +
  xlab("month of life") + guides(col = FALSE) +
  theme_bw() + ylab("Factor value") + ggtitle("MEFISTO - Factor 2")

# weights
gg_w2 <- gg_w2 + ggtitle("Weights for MEFISTO Factor 2")
```


### Figure
```{r}
cowplot::plot_grid(gg_CTF_Z2 + guides(col=F) + xlab("month of life"),
  gg_w2_CTF + theme(legend.position = "right"),
  gg_Z2,
  gg_w2+ theme(legend.position = "right"),
  labels = LETTERS[1:4], rel_widths = c(1,1.5))
```

## Correlation
```{r}
factors_CTF %>% select(Factor1, Factor2) %>%
  cor() %>% corrplot::corrplot()
get_factors(MOFAobject) %>% Reduce(rbind, .) %>%
  cor()%>% corrplot::corrplot()
```

# SessionInfo
```{r}
sessionInfo()
```
