---
title: "Evaluate simulation results"
author: "Britta Velten"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(cowplot)
library(magrittr)
options(stringsAsFactors = FALSE)

# define colors for Methods
cols2methods <- c(rep(c("steelblue3", "cyan3"),2),
                  "gray", "darkolivegreen")
names(cols2methods) <- c("MEFISTO", "MEFISTO+align",
                         "no_alignment", "alignment",
                         "MOFA2", "MEFISTO_sparse") 

# set output directory for  figures
today <- Sys.Date()
figdir = paste0("figures_simulation-", today, "/grid/")
if(!dir.exists(figdir)) {
  dir.create(figdir)
}
knitr::opts_chunk$set(dev = c("png", "pdf"), fig.path = figdir,
                      warning = FALSE, message = FALSE)

# source helper functions
source("utils_simulation.R")
```

# Load simulation results
Load results from `grid.py`.
```{r}
res <- read.csv("out/simulation_results.csv",
                sep = ',')[,-1]

# subset evaluation to available experiments
show_align <- "align" %in% unique(res$note)
show_sparse <- "sparse" %in% unique(res$note)
show_base <- "vary_N" %in% unique(res$note)

# some renaming etc
res %<>% rename(noise = noise_level)
res %<>% mutate(time = as.numeric(time))
res %<>% mutate(factor_r2 = as.numeric(ifelse(is.na(factor_r2), 0, factor_r2))) # if incorrect number of factors, set R2 to zero
res %<>% mutate(missing = as.numeric(missing) * 100) # turn to percentage

# extract base parameters of the grid
base_settings <- res %>% filter(note == "vary_N") %>%
  select(G, missing, noise) %>% unique()
base_settings <- c(base_settings, res %>%
                     filter(note == "vary_G") %>%
                     select(N) %>% unique())
df_base_settings <- data.frame(base_value = unlist(base_settings),
                               param = names(base_settings))
df_base_settings <- make_nice_names(df_base_settings)
```


# Base grid
Gather experiments and rename to nicer names.
```{r, warning=TRUE, message=TRUE, fig.width=8,fig.height=4}
# gather parameters that were varied
res_base <- res %>% gather(key = "param", value = "value",
                               c("N", "G", "noise", "missing"))
res_base %<>% filter(paste0("vary_",param) == note)

# nice names for parameters
res_base <- make_nice_names(res_base)
```

## MSE and factor recovery 
```{r main,fig.width=10, fig.height=2.2, eval = show_base}
# Plot factor recovery
pC <- res_base %>% filter(param != "Noise variance") %>% 
  ggplot(aes(x= value, y= factor_r2, col = method )) +
  stat_summary(fun.data = "mean_se", geom = "line") +
  geom_vline(data = filter(df_base_settings, param != "Noise variance"),
             aes(xintercept = base_value), lty = "dashed", col = "black", alpha =0.2) +
  facet_wrap(~param, scales = "free_x", nrow = 1 ) +
  stat_summary(fun.data = "mean_se", size = 0.1)  +
  ylab(expression("Factor recovery ("~R^{2}~")")) +
  theme_bw() + scale_color_manual(values = cols2methods)+
  xlab("") + guides(col = FALSE) +
  scale_x_continuous(breaks = scales::pretty_breaks(4), limits = c(NA, NA)) 


# Plot imputation MSE
pD <- res_base %>% filter(param != "Noise variance", !(param == "% missing samples" & value == 0)) %>% 
  ggplot(aes(x= value, y= mse, col = method )) +
  stat_summary(fun.data = "mean_se", geom = "line") +
  geom_vline(data = filter(df_base_settings, param != "Noise variance"),
             aes(xintercept = base_value), lty = "dashed", col = "black", alpha =0.2) +
  facet_wrap(~param, scales = "free_x") +
  stat_summary(fun.data = "mean_se", size = 0.1)  +
  ylab("Imputation MSE") +
  theme_bw() + scale_color_manual(values = cols2methods)+
  scale_x_continuous(breaks = scales::pretty_breaks(4), limits = c(NA, NA)) +
  xlab("") + guides(col = FALSE)

cowplot::plot_grid(pC, pD, scale = 1, rel_widths = c(1,0.95))
```


## Smoothness & sharedness
```{r}
gg_smooth <-  plot_smoothness(res_base, methods = "MEFISTO") 

gg_shared <- plot_sharedness(res_base, methods = "MEFISTO") 
```

```{r smooth_shared, eval = TRUE, fig.width=5,fig.height=6, eval = show_sparse}
cowplot::plot_grid(gg_smooth, gg_shared, labels = c("A", "B"), ncol = 1)
```


# Evaluation of using sparse Gaussian processes
```{r}
res_sparse <- res %>%
  filter(note == "sparse", method %in% c("MEFISTO", "MEFISTO_sparse")) %>% 
  gather(key = "param", value = "value", "N") 
```

## Factor recovery
```{r, eval = TRUE, fig.width=8,fig.height=4, eval = show_sparse}
# Plot factor recovery
min_r2 <- min(res_sparse$factor_r2)
miny <- min(min_r2, 0.5)
ggr2 <- ggplot(res_sparse, aes(x= value, y= factor_r2, col = method )) +
  stat_summary(fun.data = "mean_se", geom = "line") +
  stat_summary(fun.data = "mean_se", size = 0.1)  +
  ylab(expression("Factor recovery ("~R^{2}~")")) +
  xlab("Number of timepoints") + theme_bw() + ylim(c(miny,1))+
  scale_color_manual(values = cols2methods)

# Plot imputation MSE
ggmse <- ggplot(res_sparse, aes(x= value, y= mse, col = method )) +
  stat_summary(fun.data = "mean_se",geom = "line") +
  stat_summary(fun.data = "mean_se", size = 0.1) +
  ylab("MSE of imputation") +
  xlab("Number of timepoints") + theme_bw() +
  scale_color_manual(values = cols2methods)
```

## Time and Memory
```{r, fig.width=8,fig.height=4, eval = show_sparse}
# Plot time and memory
gg_time <- res_sparse %>%
  mutate(time = as.numeric(time) / 60) %>%
  ggplot(aes(x= value, y= time, col = method)) +
  stat_summary(fun.data = "mean_se",geom = "line") +
  stat_summary(fun.data = "mean_se", size = 0.1) +
  theme_bw() +
  xlab("Number of timepoints") + ylab("time [min]") +
  scale_color_manual(values = cols2methods)

gg_mem <- res_sparse %>%
  mutate(mem_usage = as.numeric(mem_usage) / 2^10 / 2^10) %>%
  ggplot(aes(x= value, y= mem_usage, col = method)) +
  stat_summary(fun.data = "mean_se",geom = "line") +
  stat_summary(fun.data = "mean_se", size = 0.1) +
  theme_bw() +
  xlab("Number of timepoints") + ylab("memory usage [MB]") +
  scale_color_manual(values = cols2methods) 
```

## Smoothness
```{r, warning=TRUE, message=TRUE, eval = show_sparse}
gg_smoothness <- plot_smoothness(res_sparse, methods = c("MEFISTO", "MEFISTO_sparse"))
```

## Joint plot for sparse GP
```{r sparse, fig.width=10,fig.height=2.5, eval = show_sparse}
gg_sparse <- cowplot::plot_grid(ggr2 + guides(col = F),
                   ggmse + guides(col = F),
                   gg_time + guides(col = F),
                   gg_mem + guides(col = F),
                   cowplot::get_legend(gg_time),
                   labels = c("A","B", "C", "D",""),
                   ncol = 5, scale = 0.98
                   )

gg_sparse
```


# Alignment evaluation
```{r}
res_align <- filter(res, note == "align") %>%
    gather(key = "param", value = "value", "N") 
```

## Factor revocery and MSE of imputation
```{r, eval = show_align, fig.width=8, fig.height=4}
# Plot factor recovery
ggr2 <- ggplot(res_align, aes(x= value, y= factor_r2,
                              col = method, lty = method == "MEFISTO+align" )) +
  stat_summary(fun.data = "mean_se", geom = "line") +
  stat_summary(fun.data = "mean_se", size = 0.1) +
  ylab(expression("Factor recovery ("~R^{2}~")")) + theme_bw() +
  scale_color_manual(values = cols2methods) +
  guides(lty = FALSE) + xlab("Number of timepoints")+
  scale_x_continuous(breaks = scales::pretty_breaks(4), limits = c(NA, NA))

# Plot imputation MSE
ggmse <- ggplot(res_align, aes(x= value, y= mse,
                               col = method, lty = method == "MEFISTO+align")) +
  stat_summary(fun.data = "mean_se", geom = "line") +
  stat_summary(fun.data = "mean_se", size = 0.1)  +
  ylab("Mean squared error of imputation") +
  theme_bw() + scale_color_manual(values = cols2methods) +
  guides(lty = FALSE)  + xlab("Number of timepoints")+
  scale_x_continuous(breaks = scales::pretty_breaks(4), limits = c(NA, NA))

```

## Smoothness & sharedness
```{r, eval = show_align}
gg_smoothness <- plot_smoothness(res_align, methods = c("MEFISTO", "MEFISTO+align")) 
gg_sharedness <- plot_sharedness(res_align, methods = c("MEFISTO", "MEFISTO+align"))
```

## Warping function MSE
```{r, fig.width=4, fig.height=3, eval=show_align}
gg_warpmse <- res_align %>%
  filter(method %in% c("MEFISTO", "MEFISTO+align")) %>%
  mutate(method = ifelse(method == "MEFISTO", "no_alignment", "alignment")) %>%
  ggplot(aes(x= value, y= warp_mse,
             col = method, lty = method == "MEFISTO+align" )) +
  stat_summary(fun.data = "mean_se", geom = "line") +
  stat_summary(fun.data = "mean_se", size = 0.1) +
  ylab("MSE of warping") + theme_bw() +
  scale_color_manual(values = cols2methods) +
  guides(lty = FALSE)  + xlab("Number of timepoints")+
  scale_x_continuous(breaks = scales::pretty_breaks(4), limits = c(NA, NA))
```

## Joint plot
```{r align, fig.width=10, fig.height=10, eval=show_align}
plot_grid(ggr2 + guides(col = F),
                   ggmse + guides(col = F),
                  get_legend(ggmse),
                    gg_smoothness+ guides(col = F, lty = FALSE),
                   gg_sharedness+ guides(col = F, lty = FALSE),
                  plot_grid(get_legend(gg_smoothness + guides(lty = FALSE)), get_legend(gg_sharedness + theme(legend.title = element_blank())), ncol =1),
                  gg_warpmse + guides(col = F),
                  get_legend(gg_warpmse),
                  labels = c("A","B", "", "C", "D", "", "E"),
                  nrow = 3)
```

# SessionInfo
```{r}
sessionInfo()
```

